Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-21 products/manuscript/Manuscript.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

One indicator of the level of virus activity present in the mosquito
population can be calculated from the information contained in dataset.
The minimum infection rate (MIR) can be calculated at various temporal
and spatial scales from trap site over one day to all sites over an
entire season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ MIR provides an estimate of the prevalence of a specific pathogen
within a mosquito population. While it does not directly measure the
risk to humans, a higher MIR in mosquitoes that commonly bite humans can
indicate a higher risk of transmission. A study pubulihsed in 2007
indicated that in years with human cases in Massachusetts between 1979
and 2004, MIR was ≥1.0 by week 31 [@hachiyaHumanEasternEquine2007]. MIR
allows for the comparison of infection rates across different
geographical areas or time periods. This is crucial for identifying
areas or times of higher risk and for evaluating the effectiveness of
control measures.

@@

*weather data is currently being reprocessed because I have decided to
switch from daily summaries to hourly summaries. new analysis will be
added in version 5* Weather data played a significant role in this phase
of the analysis. Hourly weather summaries were downloaded from NOAA's
Climate Data Portal. The dataset covers the years 2007-2023 at the
Taunton Airport in Bristol County.

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold or degree-days. Degree-days is a measure of the amount of heat
accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006].

In addition to the weather summaries, drought data will be included in
the models. Drought data was obtained from the US Drought Monitor and
the Federal Emergency Management Agency (FEMA). The data includes weekly
drought severity ratings for the entire United States. The data is
available at the county level and will be used to calculate the number
of weeks in drought conditions and the severity of the drought. For this
project, drought data for Bristol County, MA, will be used and
incorporated with regional drought indicators.

@@

*This section is under development and will be completed after analysis
is complete. The following is a general outline of the methods.*

Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
average, the United States records only about 9 cases of EEE annually.
Notably, Massachusetts reports more cases than any other state except
Florida, highlighting regional disparities in disease occurrence.
However, the incidence rate for neuroinvasive EEE cases in Bristol
County, Massachusetts, is estimated to be only around 0.06 per 100,000
people
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
low incidence of human cases poses significant challenges for
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

As stated earlier, MIR will be the main outcome of interest for this
project.

@@

*Culiseta melanura* (MEL) plays an important role in EEE virus
amplification in the enzootic cycle. Understanding the population
dynamics might provide more insight into what is driving the 2019 spike.

@@

The first models explored the relationship between virus isolation in
mosquitoes and weather conditions (average temperature by week number
and total precipitation by week). The outcome was whether there was at
least 1 positive pool in a week. The predictor in the first model was
the average maximum daily temperature the week before the pool week. The
logistic model suggest that there was a relationship between the two.
(The R code for the model can be found in the EDA-Code file:
logistic\_.pos_temp_precip.R) The average temperature in the week before
is positively associated with the likelihood of a positive result, with
an odds ratio of approximately 1.069 for each unit increase in
temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001), indicating
that higher temperatures significantly increase the probability of a
positive outcome. The model's improvement is highlighted by a decrease
in residual deviance to 405.79 from the null deviance of 417.24, upon
incorporating Lagged_Avg_TMAX as a predictor, demonstrating a better fit
to the observed data. However, with a residual deviance of 405.79 on 299
degrees of freedom, it is evident that not all variability in the
outcome is explained, suggesting the presence of other influential
factors. The model's AIC of 409.79 indicates a balance between model
complexity and predictive power, supporting its relative utility in
understanding the data. However, this relationship should be examined
more closely as temperature is likely auto correlated with progressing
weeks during summer months. While precipitation has been demonstrated to
influence the population dynamics it did not show to have any effect on
a positive result.

Another model assessed whether the temperature precipitation in April,
May, and June of each year effected the date of the first positive pool
test. The date of the first positive pool was identified and assigned a
day number from January 1st of that year. The predictors in this linear
model were the average maximum temperature for April, May, and June and
the sum of precipitation for the same time period. None of these
predictors were significant.

@@

More models assessing the effects of weather and other factors on
positive pools by mosquito species and MIR values will be developed in
Part 4 of the assignment. Future models will examine the the effects of
weather conditions on mosquito population dynamics.

## Statistical analysis

*UPDATE for Pt 4* *a better source for weather data was found and it
being processed.* *the analysis will be updated in the next version of
the document*

![](images/correlation_plot.png){fig-align="center"}

# Results

*updated in later phases of the project*

## Full analysis

*These will be generated in Pt 5* 
# Discussion 
*this section will be
completed when analysis is done*

@@

*this section will be updated after analysis is complete* The accuracy
and reliability of the mosquito collection data is excellent. The same
procedures and sentinel sites have were in place for the entire
analytical horizon. Further, there were only two people responsible for
sorting, identifying and counting species over the entire period.

MIR is not a perfect measure of virus activity in mosquitoes. It tends
to underestimate the level of virus circulating in the environment.
However, other measures require equal pool numbers for testing. There is
a correction factor for variable pool numbers and I will explore this
option in later phases of this project.

@@

## Conclusions *this section will be completed when analysis is done*

>>>>

One indicator of the level of virus activity present in the mosquito
population can be calculated from the information contained in the
mosquito dataset. The minimum infection rate (MIR) can be calculated at
various temporal and spatial scales from trap site over one day to all
sites over an entire season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ MIR provides an estimate of the prevalence of a specific pathogen
within a mosquito population. While it does not directly measure the
risk to humans, a higher MIR in mosquitoes that commonly bite humans can
indicate a higher risk of transmission. A study pubulihsed in 2007
indicated that in years with human cases in Massachusetts between 1979
and 2004, MIR was ≥1.0 by week 31 [@hachiyaHumanEasternEquine2007]. MIR
allows for the comparison of infection rates across different
geographical areas or time periods. This is crucial for identifying
areas or times of higher risk and for evaluating the effectiveness of
control measures.

@@

Weather plays a significant role in the lifecyle of the mosquito and
influences the rate at which a pathogen can develop in a mosquito.
Hourly and weather summaries were downloaded from NOAA's Climate Data
Portal. The datasets cover the years 2007-2023 at the Taunton Airport in
Bristol County.

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold or degree-days. Degree-days is a measure of the amount of heat
accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006]. To simplify this
variable, the number of hours the temperature was ≥50°F and ≤86°F. These
thresholds were chosen based on the known temperature range for mosquito
development and virus transmission.

In addition to the weather summaries, drought data will be included in
the models. West Nile Virus (WNV) research suggests that drought has
been a driver in WNV outbreaks. While the lack of water can lead to
smaller mosquito populations, drought can lead to higher prevelance of
mosquito infection [@paullDroughtImmunityDetermine2017]. Drought data
was obtained from the US Drought Monitor and the Federal Emergency
Management Agency (FEMA). The data includes weekly drought severity
ratings for the entire United States. The data is available at county
and regional levels and will be used to calculate the number of weeks in
drought conditions and the severity of the drought. For this project,
drought data for Bristol County, MA, will be used and incorporated with
FEMA East Coast regions (I-IV) drought indicators.

@@

Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
average, the United States records only about 9 cases of EEE annually.
Notably, Massachusetts reports more cases than any other state except
Florida, highlighting regional disparities in disease occurrence.
However, the incidence rate for neuroinvasive EEE cases in Bristol
County, Massachusetts, is estimated to be only around 0.06 per 100,000
people
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
low incidence of human cases poses significant challenges for using
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

MIR will be the main outcome of interest for this project. MIR values
will be calculated for each week during the surveillance season over the
entire county. To reduce bias, only mosquitoes captured using carbon
dioxide-baited CDC light traps will be used to to calculate MIR. These
traps are the standard EEE surveillance and are used to monitor the
actively feeding mosquito population. Additionally, these trap types
have been used in this area for over 40 years, many at the same sites
currently in use.

After MIR was calculated and joined with weather and climate data, the
distribution of the MIR was examined more closely. About 72% of the
records were zero values, which is not uncommon for this type of data.
The non-zero values were skewed to the left and there was strong
evidence of over-dispersion.

![](images/mir_density_plot.png){fig-align="center"}

The average MIR was 1.26, with an observed variance of 7.56, resulting
in a Variance-to-Mean Ratio (VMR) of 6.00. This pronounced
overdispersion suggests that mosquito infection rates were not uniformly
distributed over time or space. Instead, they may have been influenced
by a range of factors, such as environmental conditions or variations in
amplifying host abundance.

The data was visualized to identify trends and relationships between the
variables. There were no clear patterns in the data between MIR and the
potential predictors. However, several varibles showed a non-linear
relationship with MIR. For example, the relationship between MIR and
average temperature was not linear. Instead, MIR increased with
temperature up to a certain point, then decreased. This suggests that
the relationship between temperature and MIR may be more complex than a
simple linear model can capture.

In response to the observed non-linear relationships in the data, a
two-stage Generalized Additive Model (GAM) analysis was used. GAMs
provide a versatile framework for capturing complex, non-linear
relationships between response variables and multiple predictors, which
are often encountered in ecological data sets.

In the first stage of the analysis, the GAM was utilized to address the
high incidence of zeros in the dataset—an indicator of overdispersion.
This zero-inflated model allowed us to separately model the presence
versus absence of the outcome variable, thereby appropriately accounting
for the excessive zeros.

Subsequently, the second stage of the GAM focused on modeling the
continuous part of the MIR data, relating it to the suite of potential
predictors. This step incorporated smooth functions of the predictors,
permitting the data to inform the shape of the relationship without
imposing a predetermined form, as would be the case in a traditional
linear model.

While the two-stage GAM appeared to be a good fit, there were too few
observations to provide a robust estimate of the relationship between
the predictors and MIR. The original dataset contained over 40,000
observations, but after reformatting the data to fit a rectangular shape
needed for analysis, there were less than 400 observations in the final
set. After the zeros were removed for part 2 of the GAM, the sample
number of observations left was less than 100. Because of this, it was
decided to use a single GAM with a dataset that included zero and
non-zero observations. Additionally, all trap types were included in the
analysis, not just the carbon dioxide-baited CDC light traps. While some
precision was lost when the one part model was adopted, the gain in
observations was deemed more valuable.

@@

*Culiseta melanura* (MEL) plays an important role in EEE virus
amplification in the enzootic cycle. Understanding the population
dynamics might provide more insight into what is driving the spikes,
like in 2019.

@@

The first models explored the relationship between virus isolation in
mosquitoes and weather conditions (average temperature by week number
and total precipitation by week). The outcome was whether there was at
least 1 positive test for pool in a given week. The predictor in the
first model was the average maximum daily temperature the week before
the mosquito collection week. The logistic model suggests that there was
a relationship between the two. (The R code for the model can be found
in the analysis-code file: analysis.qmd) The average temperature in the
week before is positively associated with the likelihood of a positive
result, with an odds ratio of approximately 1.069 for each unit increase
in temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001),
indicating that higher temperatures significantly increase the
probability of a positive outcome. The model's improvement is
highlighted by a decrease in residual deviance to 405.79 from the null
deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a predictor,
demonstrating a better fit to the observed data. However, with a
residual deviance of 405.79 on 299 degrees of freedom, it is evident
that not all variability in the outcome is explained, suggesting the
presence of other influential factors. The model's AIC of 411.59
indicates a balance between model complexity and predictive power,
supporting its relative utility in understanding the data. However, this
relationship should be examined more closely as temperature is likely
auto correlated with progressing weeks during summer months. While
precipitation has been demonstrated to influence the population dynamics
it did not show to have any effect on a positive result. This was
initailly surprising, but upon after considering the specific habitat of
the main EEE enzootic vector, *Culiseta melanura*, it makes sense. MEL
larvae develop in well insulated crypts that are less sensitive
evaporative processes than surface water. The effect of precipitation on
MEL larval habitat abundance may be less pronounced than other vector
species that develop in open, standing water, like ditches and
containers. Another model assessed whether the temperature precipitation
in April, May, and June of each year effected the date of the first
positive pool test (R file: PT3_edacode). The date of the first positive
pool was identified and assigned a day number from January 1st of that
year. The predictors in this linear model were the average maximum
temperature for April, May, and June and the sum of precipitation for
the same time period. None of these predictors were significant.

@@

## Statistical analysis
Before any models were assessed, the RSME for the null model was calculated (2.40). This was used as the basis for assessing model improvement.

Analysis began with a Generalized Additive Model (GAM) with predictors that were selected based on the exploratory single variable GAM. The variables were chosen if their p=values were <0.05. Eight variables met that condition including: the number of hours below 50°F the week before the mosquito collection week (Hrs_Below_50_lag), the number of hours below 50°F the week of collection (Hrs_Below_50_current), the drought severity index for East Coast FEMA regions (DSCI_fema), the average hourly relative humidity the week of collection (Avg_HourlyRelativeHumidity_current), the average hourly windspeed the week of collection (Avg_HourlyWindSpeed_current), the average number of daylight hours per day (average_daylight_hours), the average number of MEL captured by trap (Average_MEL), and the proportion of MEL captured out of the total number of mosquitoes trapped (MEL_Prop).

All models included training (.75) and a testing (.25) dataset that were generated using the random seed 123. All analysis can be found in the analysis.qmd file.  The first model was a GAM with all eight predictors, followed by a Random Forest, LASSO, Generalized Additive Model with Sparsity (GAMS), and a Gradient Boosting Machine (GBM) model.

As expected, the RSME is lower for the training dataset than the testing dataset for all models. The RMSE values are displayed in the table below.

```{r}
# Define the file path using here()
file_path_rds <- here("results", "tables", "rmse_table.rds")

# Read the RDS file
rmse_table <- readRDS(file_path_rds)
rmse_table
```
# Discussion
The null model RMSE was 2.40. All models showed improvement over the null model on both the training and testing datasets. The GAM shows significant improvement over the null model for the training set, but only slightly improves over the null for the testing set. This suggests that beyond capturing the mean response, this GAM is not well generalized. The substantial improvement in the training set RSME for GBM and Random Forest coupled with less improvement over the null for the testing set, suggest a tendency for overfitting. These models may be capturing the training data's patterns that do not gerneralize well to unseen data. The the gap between the training and testing RMSE for the LASSO and GAMS models suggest there is less overfitting and are slightly better at generalizing than the others. However, given the non-linear relationships, the LASSO model is likely not appropriate for this analysis.

@@

The accuracy
and reliability of the mosquito collection data is excellent. The same
procedures and sentinel sites have were in place for the entire
analytical horizon. Further, there were only two people responsible for
sorting, identifying and counting species over the entire period. However, the data is not without limitations. Mosquito surveillance is inherently biased. Traps are set in locations that are most likely to capture infected mosquitoes. While this is a positive for early warning of virus activity, it is not necessarily representative of the surrounding area. Probably the biggest limitation is the small number of observations. The mosquito season is short in Massachusetts and funding for trapping and testing mosquitoes is limited. Mosquito control efforts, like pesticide applications, are missing from this analysis. These efforts are designed to lower the levels of virus transmitting mosquitoes and would likely provide more insight into the relationship between the predictors and virus activity.

MIR is not a perfect measure of virus activity in mosquitoes. It tends
to underestimate the level of virus circulating in the environment.
However, other measures require equal pool numbers for testing. There is
a correction factor for variable pool numbers and I will explore this
option in later phases of this project. Ideally, the outcome would be human or animal cases. However, these cases are rare and modeling them given the limited amount of surveillance data may not be possible.

Initially, the lack of significance for most temperature and precipitation variable was surprising. However, upon closer examination, it makes sense. The main enzootic vector of EEE and the one targeted for surveillance, *Culiseta melanura*, develops in well insulated crypts that are less sensitive to ambient air temperature changes. The effect of temperature changes on MEL development speed may be less pronounced than other vector species that develop in open, standing water, like ditches and containers. Precipitation increases the amount of standing water and increases the chances of larval survival during the mosquito season and also during the winter months for those species that overwinter in the larval stage. For EEE, the main enzootic vector, MEL, overwinters as larvae.

@@

Likely the biggest limitation 
## Conclusions 
While all models demonstrated improvement over null model, none of them were robust enough to be used for reliable prediction. In the next phases of this project, the data will be expanded to include more years and cover a larger portion of Massachusetts. In additon to increasing the the number of observations, other sources of data will be incorporated. Including virus activity from surrounding regions and areas from which birds migrate to Massachusetts.
